Component response rate variation drives stability in large complex systems

### 2018-05-20 ms.Rmd
Update small tweaks to ms, mainly instructions for making a TEX file to submit to arxiv

---
title: "Component response rate variation drives stability in large complex systems"
author: "Brad Duthie"
bibliography: references.bib
output:
  html_document: default
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "Brad Duthie"
bibliography: references.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

<!---
"This will generate the intermediate files that are actually used to generate the PDF. You need only go to the directory where your .Rmd file is, gather the .tex and image directory (called filename_files) with"

arxiv comand: tar cvfz filename.tgz filename.tex filename_files
--->

### 2018-05-20 ms.tex
Update small tweaks to ms, mainly instructions for making a TEX file to submit to arxiv

\textbf{The stability of a complex system generally decreases with
increasing system size, as is demonstrated by random matrix
theory\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
This counter-intuitive result, first shown by
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}, is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{3},\protect\hyperlink{ref-Gray2009}{4}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{5},\protect\hyperlink{ref-MacArthur2010}{6}}
and
socio-economic\textsuperscript{\protect\hyperlink{ref-Haldane2011}{7}--\protect\hyperlink{ref-Bardoscia2017}{9}}
networks. Much attention has especially been given to the stability of
ecological communities such as food webs or mutualist networks, with
recent work investigating how different community structures affect
stability\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Mougi2012}{10}--\protect\hyperlink{ref-Patel2018}{14}}.
But more broadly, stabilising mechanisms in complex systems remain
under-developed, and the effect of variation in the response rate of
individual system components remains an open
problem\textsuperscript{\protect\hyperlink{ref-Allesina2015}{15}}. Here
I show that when components of a complex system respond to system
dynamics at different rates (\(\boldsymbol{\omega}\)), the potential for
system stability is markedly increased. Stability increases due to the
clustering of some eigenvalues toward the centre of eigenvalue
distributions despite the destabilising effect of higher variation among
interaction strengths (\(\boldsymbol{\sigma^{2}}\)). This effect of
variation in \(\boldsymbol{\omega}\) becomes increasingly important as
system size increases, to the extent that the largest stable complex
systems would otherwise be unstable if not for
\(\boldsymbol{Var(\omega)}\). My results therefore reveal a previously
unconsidered driver of system stability that is likely to be pervasive
across all complex systems. Future research in complex systems should
therefore account for the varying response rates of individual system
components when assessing whole system stability.}

@@

The above stability criterion assumes that individual components respond
to perturbations of the system at the same rate (\(\omega\)), but this
is highly unlikely in any complex system. In ecological communities, for
example, the rate at which population density changes following
perturbation will depend on the generation time of individuals, which
might vary by orders of magnitude among species. Species with short
generation times will respond quickly (high \(\omega\)) to perturbations
relative to species with long generation times (low \(\omega\)).
Similarly, the speed at which individual banks respond to perturbations
in financial networks, or individuals or institutions respond to
perturbations in complex social networks, is likely to vary. The effect
of such variance has not been investigated in complex systems theory.
Intuitively, variation in \(\omega\) might be expected to decrease
system stability by introducing a new source of variation into the
system and thereby increasing \(\sigma\). Here I show why, despite
higher \(\sigma\), complex systems in which \(\omega\) varies actually
tend to be more stable, especially when \(S\) is high.

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a scalar value
\(\omega_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\omega\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\omega_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\omega_{i} = 1.95\) and half of
\(\omega_{i} = 0.05\). This models one system in which \(\omega\) is
invariant and one in which \(\omega\) varies, but systems are otherwise
identical (note \(E[\omega_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by \(y_{i}\) to
generate \(M\). When \(\omega_{i} = 1\), eigenvalues of \(M\) are
distributed uniformly within a circle centred at (\(-1, 0\)) with a
radius of 1.265 (Fig. 1a). Hence, the real components of eigenvalues are
highly unlikely to all be negative when all \(\omega_{i} = 1\). But when
\(\omega_{i}\) values are separated into two groups, eigenvalues are no
longer uniformly distributed (Fig. 1b). Instead, two distinct clusters
of eigenvalues appear (grey circles in Fig. 1b), one centred at
(\(-1.95, 0\)) and the other centred at (\(-0.05, 0\)). The former has a
large radius, but the real components have shifted to the left (in
comparison to when \(\omega = 1\)) and all \(\Re({\lambda}) < 0\). The
latter cluster has real components that have shifted to the right, but
has a smaller radius. Overall, for 1 million randomly assembled \(M\),
this division between slow and fast component response rates results in
more stable systems: 0 stable given \(\omega = 1\) versus 0 stable given
\(\omega = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\omega\) can be observed
by sampling \(\omega_{i}\) values from various distributions. I now
focus on a uniform distribution where \(\omega \sim \mathcal{U}(0, 2)\)
(see Supporting Information for other distributions, which give similar
results). As with the case of \(\omega = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\omega] = 1\) when \(\omega \sim \mathcal{U}(0, 2)\), allowing
comparison of \(M\) before and after variation in component response
rate. Figure 2 shows a comparison of eigenvalue distributions given
\(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{17}}, when
\(\omega = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 12.649.
Uniform variation in \(\omega\) leads to a non-uniform distribution of
eigenvalues, some of which are clustered tightly around the centre of
the distribution, but others of which are spread outside the former
radius of 12.649 (red circle Fig 2b). This larger radius occurs because
the addition of \(Var(\omega)\) increases the realised \(\sigma\) of
\(M\). The clustering and spreading of eigenvalues introduced by
\(Var(\omega)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\omega = 1\), the effect of \(Var(\omega)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold of \(\sigma\sqrt{SC} > 1\).

To investigate the effect of \(Var(\omega)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from \(2-32\) (see Supporting Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\omega = 1\)
versus \(\omega \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\omega = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\omega)\) than when \(\omega = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\omega)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\omega)\) would not have been stable given \(\omega = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\omega)\) versus \(27\) given \(\omega = 1\) (see Supporting
Information for full results). This suggests that the stability of large
systems might be dependent upon variation in the response rate of their
individual components, meaning that factors such as generation time (in
ecological networks), transaction speed (in economic networks), or
communication speed (in social networks) needs to be considered when
investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise
that \(Var(\omega)\) is not stabilising per se; that is, adding
variation in \(\omega\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supporting Information). Rather, systems that are observed to be stable
are more likely to vary in \(\omega\), and for this \(Var(\omega)\) to
be critical to their stability. This is caused by the shift in the
distribution of eigenvalues that occurs by introducing \(Var(\omega)\)
(Fig. 1b, 2b), which can sometimes result in all \(\Re(\lambda) < 0\)
but might also increase \(\Re(\lambda)\) values.

@@

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\omega} = 1.95\)) and slow
(\(\boldsymbol{\omega} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\omega] = 1\) (i.e., only the
distribution of \(\omega\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\omega = 1\); distributions of points are
uniformly distributed within the grey circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\omega = 1.95\) and half
\(\omega = 0.05\); distributions of points can be partitioned into one
large circle of \(\sigma\sqrt{SC} =\) 1.718 centred at
\(\omega = -1.95\) and one small circle of \(\sigma\sqrt{SC} =\) 0.044
centred at \(\omega = -0.05\). In a, the maximum real eigenvalaue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 0 were stable when
\(\omega = 1\) while 0 were stable when \(\omega = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\omega}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\omega)\) where
\(E[\omega] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\omega \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.55 and 0.554 for a and b, respectively.

@@

\textbf{Figure 3: Stability of large complex systems with and without
variation in component response rate (\(\boldsymbol{\omega}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S\)) given \(C = 1\), and the proportion of systems in which
variation in \(\omega\) is critical for system stability. For each
\(S\), 1 million complex systems are randomly generated. Stability of
each complex system is tested given variation in \(\omega\) by randomly
sampling \(\omega \sim \mathcal{U}(0, 2)\). Stability given
\(Var(\omega)\) is then compared to stability in an otherwise identical
system in which \(\omega = E[\mathcal{U}(0, 2)]\) for all components.
Light and dark grey bars show the number of stable systems in the
absence and presence of variance in \(\omega\), respectively. The black
line shows the proportion of systems that are stable when
\(Var(\omega) > 0\), but would be unstable if \(Var(\omega) = 0\).

>>>>

\textbf{The stability of a complex system generally decreases with
increasing system size, as is demonstrated by random matrix
theory\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
This counter-intuitive result, first shown by
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}, is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{3},\protect\hyperlink{ref-Gray2009}{4}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{5},\protect\hyperlink{ref-MacArthur2010}{6}}
and
socio-economic\textsuperscript{\protect\hyperlink{ref-Haldane2011}{7}--\protect\hyperlink{ref-Bardoscia2017}{9}}
networks. Much attention has especially been given to the stability of
ecological communities such as food webs or mutualist networks, with
recent work investigating how different community structures affect
stability\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Mougi2012}{10}--\protect\hyperlink{ref-Patel2018}{14}}.
But more broadly, stabilising mechanisms in complex systems remain
under-developed, and the effect of variation in the response rate of
individual system components remains an open
problem\textsuperscript{\protect\hyperlink{ref-Allesina2015}{15}}. Here
I show that when components of a complex system respond to system
dynamics at different rates (\(\boldsymbol{\gamma}\)), the potential for
system stability is markedly increased. Stability increases due to the
clustering of some eigenvalues toward the centre of eigenvalue
distributions despite the destabilising effect of higher variation among
interaction strengths (\(\boldsymbol{\sigma^{2}}\)). This effect of
variation in \(\boldsymbol{\gamma}\) becomes increasingly important as
system size increases, to the extent that the largest stable complex
systems would otherwise be unstable if not for
\(\boldsymbol{Var(\gamma)}\). My results therefore reveal a previously
unconsidered driver of system stability that is likely to be pervasive
across all complex systems. Future research in complex systems should
therefore account for the varying response rates of individual system
components when assessing whole system stability.}

@@

The above stability criterion assumes that individual components respond
to perturbations of the system at the same rate (\(\gamma\)), but this
is highly unlikely in any complex system. In ecological communities, for
example, the rate at which population density changes following
perturbation will depend on the generation time of individuals, which
might vary by orders of magnitude among species. Species with short
generation times will respond quickly (high \(\gamma\)) to perturbations
relative to species with long generation times (low \(\gamma\)).
Similarly, the speed at which individual banks respond to perturbations
in financial networks, or individuals or institutions respond to
perturbations in complex social networks, is likely to vary. The effect
of such variance has not been investigated in complex systems theory.
Intuitively, variation in \(\gamma\) might be expected to decrease
system stability by introducing a new source of variation into the
system and thereby increasing \(\sigma\). Here I show why, despite
higher \(\sigma\), complex systems in which \(\gamma\) varies actually
tend to be more stable, especially when \(S\) is high.

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a scalar value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\gamma_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\gamma_{i} = 1.95\) and half of
\(\gamma_{i} = 0.05\). This models one system in which \(\gamma\) is
invariant and one in which \(\gamma\) varies, but systems are otherwise
identical (note \(E[\gamma_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by \(y_{i}\) to
generate \(M\). When \(\gamma_{i} = 1\), eigenvalues of \(M\) are
distributed uniformly within a circle centred at (\(-1, 0\)) with a
radius of 1.265 (Fig. 1a). Hence, the real components of eigenvalues are
highly unlikely to all be negative when all \(\gamma_{i} = 1\). But when
\(\gamma_{i}\) values are separated into two groups, eigenvalues are no
longer uniformly distributed (Fig. 1b). Instead, two distinct clusters
of eigenvalues appear (grey circles in Fig. 1b), one centred at
(\(-1.95, 0\)) and the other centred at (\(-0.05, 0\)). The former has a
large radius, but the real components have shifted to the left (in
comparison to when \(\gamma = 1\)) and all \(\Re({\lambda}) < 0\). The
latter cluster has real components that have shifted to the right, but
has a smaller radius. Overall, for 1 million randomly assembled \(M\),
this division between slow and fast component response rates results in
more stable systems: 0 stable given \(\gamma = 1\) versus 0 stable given
\(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I now
focus on a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\)
(see Supporting Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(M\) before and after variation in component response
rate. Figure 2 shows a comparison of eigenvalue distributions given
\(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{17}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 12.649.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues, some of which are clustered tightly around the centre of
the distribution, but others of which are spread outside the former
radius of 12.649 (red circle Fig 2b). This larger radius occurs because
the addition of \(Var(\gamma)\) increases the realised \(\sigma\) of
\(M\). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold of \(\sigma\sqrt{SC} > 1\).

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from \(2-32\) (see Supporting Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see Supporting
Information for full results). This suggests that the stability of large
systems might be dependent upon variation in the response rate of their
individual components, meaning that factors such as generation time (in
ecological networks), transaction speed (in economic networks), or
communication speed (in social networks) needs to be considered when
investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise
that \(Var(\gamma)\) is not stabilising per se; that is, adding
variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supporting Information). Rather, systems that are observed to be stable
are more likely to vary in \(\gamma\), and for this \(Var(\gamma)\) to
be critical to their stability. This is caused by the shift in the
distribution of eigenvalues that occurs by introducing \(Var(\gamma)\)
(Fig. 1b, 2b), which can sometimes result in all \(\Re(\lambda) < 0\)
but might also increase \(\Re(\lambda)\) values.

@@

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\gamma} = 1.95\)) and slow
(\(\boldsymbol{\gamma} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\gamma] = 1\) (i.e., only the
distribution of \(\gamma\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\gamma = 1\); distributions of points are
uniformly distributed within the grey circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\gamma = 1.95\) and half
\(\gamma = 0.05\); distributions of points can be partitioned into one
large circle of \(\sigma\sqrt{SC} =\) 1.718 centred at
\(\gamma = -1.95\) and one small circle of \(\sigma\sqrt{SC} =\) 0.044
centred at \(\gamma = -0.05\). In a, the maximum real eigenvalaue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 0 were stable when
\(\gamma = 1\) while 0 were stable when \(\gamma = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.555 and 0.549 for a and b, respectively.

@@

\textbf{Figure 3: Stability of large complex systems with and without
variation in component response rate (\(\boldsymbol{\gamma}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S\)) given \(C = 1\), and the proportion of systems in which
variation in \(\gamma\) is critical for system stability. For each
\(S\), 1 million complex systems are randomly generated. Stability of
each complex system is tested given variation in \(\gamma\) by randomly
sampling \(\gamma \sim \mathcal{U}(0, 2)\). Stability given
\(Var(\gamma)\) is then compared to stability in an otherwise identical
system in which \(\gamma = E[\mathcal{U}(0, 2)]\) for all components.
Light and dark grey bars show the number of stable systems in the
absence and presence of variance in \(\gamma\), respectively. The black
line shows the proportion of systems that are stable when
\(Var(\gamma) > 0\), but would be unstable if \(Var(\gamma) = 0\).
